%!TEX root = ../thesis.tex

% Thesis Acknowledgements ------------------------------------------------


\begin{acknowledgements}
The past three and a half years have been the most challenging, yet very rewarding experience. Quite a few people have
made it possible, and I am extremely grateful to all of them. First of all, I would like to thank my supervisor, Greg
Heath, for his valuable guidance, advice and constructive suggestions, especially in writing up this thesis. I also
thank Joel Goldstein for leading the research group and being a wise and kind host to our weekly meetings. Your advice
has been greatly appreciated.

This work would not have been possible without some serious funding. I would like to thank the Faculty of Science and
the selection panel for awarding me the University of Bristol Postgraduate Research Scholarship. It has been a great
honour and I hope I managed to live up to it. I'm also very thankful to Dave Newbold (and Joel, again) for arranging the
additional funding for my work at CERN, as otherwise my time there would have been rather miserable despite the great
privilege of witnessing the Higgs boson discovery.

Many thanks to \L{}ukasz Kreczko, Jeson Jacob and Phil Symonds for being great teammates over the last few years. I am
particularly grateful to Luke who has been a nearly endless source of knowledge, and always willing to help. K-thx-bye!

A number of people had helped me during my placement at CERN. In particular, I would like to thank Martijn Mulders and
Enrique Palencia for their collaboration and helpful advice on the top mass analysis. My work on the high-level trigger
was greatly influenced by St\'{e}phanie Beauceron and Sam Harper -- thank you for guidance and useful advice. And special
thanks to the UK Liaison Office at CERN for helping me deal with the French bureaucracy.

I would also like to express my gratitude to Alexei Safonov and Jim Pivarski for introducing me to CMS back in 2008 and
giving me a chance to appreciate the collaboration spirit, and get inspired by it. Thank you, without your help and good
will this PhD would not have been possible.

The warmest thanks go to all my friends and family, and especially to my parents. Your continuous encouragement and
support throughout all this time have been immensely appreciated. %Pardon my Russian: \foreignlanguage{russian}{Спа-си-бо}!

\end{acknowledgements}

% ------------------------------------------------------------------------

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../thesis"
%%% End: 
